---
title: "NYPD Shootings Analysis"
author: "H. Justino"
date: '2024-12-09'
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# NYPD Crime Info


### Import Data into variable
```{r, import_data_check, echo=FALSE}
file_path <- "../datasets/NYPD_Shooting_Incident_Data__Historic_.csv"
if (file.exists(file_path)) {
  nypd_data <- read.csv(file_path)
} else {
  stop("File not found: ", file_path)
}
```

### Summarize NYPD data
```{r, summarize, echo=FALSE}
summary(nypd_data)
head(nypd_data, 5)
```

#Select Columns
```{r, select_columns, warning=FALSE, echo=FALSE}
# Select relevant columns and filter out rows with NA or invalid values
library(dplyr)

nypd_data_filtered <- nypd_data %>%
  select(Longitude, Latitude, INCIDENT_KEY) %>%
  filter(!is.na(Longitude) & !is.na(Latitude) & Longitude >= -180 & Longitude <= 180 & Latitude >= -90 & Latitude <= 90)

head(nypd_data_filtered, 5)
```

# Plot points on the map

#### I want to plot where the incidents occur in the city. I used an interactive library.


```{r, year_map, echo=FALSE}
library(shiny)
library(leaflet)

# Select relevant columns and filter out rows with NA or invalid values
library(dplyr)

nypd_data_filtered_year <- nypd_data %>%
  select(Longitude, Latitude, INCIDENT_KEY, OCCUR_DATE) %>%
  filter(!is.na(Longitude) & !is.na(Latitude) & !is.na(OCCUR_DATE) & Longitude >= -180 & Longitude <= 180 & Latitude >= -90 & Latitude <= 90)


# Convert OCCUR_DATE to Date type and extract year
nypd_data_filtered_year$OCCUR_DATE <- as.Date(nypd_data_filtered_year$OCCUR_DATE, format="%m/%d/%Y")
nypd_data_filtered_year$year <- format(nypd_data_filtered_year$OCCUR_DATE, "%Y")

# Define UI for application
ui <- fluidPage(
  titlePanel("NYPD Shooting Incidents"),
  sidebarLayout(
    sidebarPanel(
      selectInput("year", "Select Year:", choices = sort(unique(nypd_data_filtered_year$year), decreasing = TRUE), selected = sort(unique(nypd_data_filtered_year$year), decreasing = TRUE)[1])
    ),
    mainPanel(
      leafletOutput("map")
    )
  )
)

# Define server logic
server <- function(input, output, session) {
  filtered_data <- reactive({
    nypd_data_filtered_year %>% filter(year == input$year)
  })
  
  output$map <- renderLeaflet({
    leaflet(filtered_data()) %>%
      addTiles() %>%
      addCircleMarkers(~Longitude, ~Latitude, popup = ~as.character(INCIDENT_KEY), color = "red", radius = 5)
  })
}

# Run the application 
shinyApp(ui = ui, server = server)
```

# Pie chart based on Victim Sex

#### Gender distribution of the victims of crime. Men are the overwhelming participants in crime.
```{r, victim_sex, warning=FALSE, echo=FALSE}
library(ggplot2)

# Summarize data for pie chart
vic_sex_summary <- nypd_data %>%
  filter(!is.na(VIC_SEX)) %>%
  group_by(VIC_SEX) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  filter(percentage >= 0.5)  # Filter out categories with less than 0.5% 

# Create pie chart
ggplot(vic_sex_summary, aes(x = "", y = count, fill = VIC_SEX)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  geom_text(aes(label = paste0(round(percentage, 1), "% (", count, ")")), 
            position = position_stack(vjust = 0.5)) +
  labs(title = "Distribution of Victim Sex", fill = "Victim Sex")
```

# Display Shooting Rate by Borough

#### Nearly 40% of all shootings in the city happens in Brooklyn.
```{r, crime_rate_by_borough, warning=FALSE, echo=FALSE}
library(ggplot2)
  # Summarize data to calculate crime rate by borough
borough_summary <- nypd_data %>%
  filter(!is.na(BORO)) %>%
  group_by(BORO) %>%
  summarise(crime_count = n()) %>%
  mutate(crime_rate = crime_count / sum(crime_count) * 100)

# Create bar chart for crime rate by borough
ggplot(borough_summary, aes(x = reorder(BORO, -crime_rate), y = crime_rate, fill = BORO)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(crime_rate, 1), "%")), vjust = -0.5) +
  labs(title = "Percentage of Crime by Borough", x = "Borough", y = "Percentage of All Crime (%)") +
  theme_minimal()
```

## Number of shootings by time of the day.

#### There is a U-Shaped pattern, where shootings decrease during work hours. They increase in the evening peaking until midnight where they slowly decline until a shar decrease at 5am.
```{r, crime_by_day, warning=FALSE, echo=FALSE}

# Select relevant columns and filter out rows with NA or invalid values
library(dplyr)

nypd_data_filtered <- nypd_data %>%
  select(OCCUR_TIME) %>%
  filter(!is.na(OCCUR_TIME))

# Convert OCCUR_TIME to a suitable format 
nypd_data_filtered$hour <- as.numeric(substr(nypd_data_filtered$OCCUR_TIME, 1, 2))

library(ggplot2)

crime_time_plot <- nypd_data_filtered %>%
  group_by(hour) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = hour, y = count)) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, 23, by = 2)) +
  labs(title = "Distribution of shootings at Different Times of the Day",
       x = "Hour of the Day",
       y = "Number of Incidents") +
  theme_minimal()

# Print the plot
print(crime_time_plot)
```


## Import NYC Census Data

### From US Census Bureau: https://www.census.gov/quickfacts/fact/table/newyorkcountynewyork,richmondcountynewyork,kingscountynewyork,queenscountynewyork,bronxcountynewyork,newyorkcitynewyork/PST045222

#### I imported NYC Census data because I want to analyze the rate at whcih shootings happen by borough. Brooklyn has an exceptionally high number of shootings, but does it match the population?
```{r, census_summary, warning=FALSE, echo=FALSE}
file_path <- "../datasets/census.csv"
if (file.exists(file_path)) {
  census_data <- read.csv(file_path)
} else {
  stop("File not found: ", file_path)
}

summary(census_data)

head(census_data, 15)
```

#### Population of the city (2023 Estimate)
```{r, population, warning=FALSE, echo=FALSE}

# Select the columns with borough names and their population estimates
population_estimates <- census_data %>%
  select(`Fact`, `New.York.County..New.York`, `Richmond.County..New.York`, `Kings.County..New.York`, `Queens.County..New.York`, `Bronx.County..New.York`, `New.York.city..New.York`) %>%
  filter(Fact == "Population estimates, July 1, 2023, (V2023)")


# Rename columns to match the names in the crime data
census_data <- census_data %>%
  rename(
    MANHATTAN = `New.York.County..New.York`,
    STATEN_ISLAND = `Richmond.County..New.York`,
    BROOKLYN = `Kings.County..New.York`,
    QUEENS = `Queens.County..New.York`,
    BRONX = `Bronx.County..New.York`,
    NYC = `New.York.city..New.York`
  )

# Select the row with population estimates for all the boroughs
population_estimates <- census_data %>%
  filter(Fact == "Population estimates, July 1, 2023, (V2023)") %>%
  select(Fact, MANHATTAN, STATEN_ISLAND, BROOKLYN, QUEENS, BRONX, NYC)

# Print the selected row
print(population_estimates)
```

### Summarize Shootings by Borough
```{r, summarize_crimes_by_borough, warning=FALSE, echo=FALSE}
library(dplyr)
library(knitr)

# Summarize data to calculate the number of shootings by borough
borough_crime_summary <- nypd_data %>%
  filter(!is.na(BORO)) %>%
  group_by(BORO) %>%
  summarise(crime_count = n()) %>%
  arrange(desc(crime_count))

# Display the summary as a table
kable(borough_crime_summary, col.names = c("Borough", "Number of Shootings"), caption = "Number of Shootings by Borough")
```


### Summarize shootings by Borough and Calculate Crime Rates
```{r, summarize_crimes_and_calculate_rates, message=FALSE, echo=FALSE}
library(dplyr)
library(knitr)

# Summarize data to calculate the number of shootings by borough
borough_crime_summary <- nypd_data %>%
  filter(!is.na(BORO)) %>%
  group_by(BORO) %>%
  summarise(crime_count = n()) %>%
  arrange(desc(crime_count))

# Convert population estimates to numeric
population_estimates <- population_estimates %>%
  mutate(across(MANHATTAN:NYC, ~ as.numeric(gsub(",", "", .))))

# Calculate crime rates per 100,000 people
crime_rates <- borough_crime_summary %>%
  mutate(
    population = case_when(
      BORO == "MANHATTAN" ~ population_estimates$MANHATTAN,
      BORO == "STATEN ISLAND" ~ population_estimates$STATEN_ISLAND,
      BORO == "BROOKLYN" ~ population_estimates$BROOKLYN,
      BORO == "QUEENS" ~ population_estimates$QUEENS,
      BORO == "BRONX" ~ population_estimates$BRONX,
      TRUE ~ NA_real_
    ),
    crime_rate_per_100k = (crime_count / population) * 100000
  )

# Display the summary as a table
kable(crime_rates, col.names = c("Borough", "Number of shootings", "Population", "Crime Rate per 100,000"), caption = "Crime Rates by Borough")

```

## Shooting Rate By Borough Graph

#### We can see the order of the crime rate by borough in the city. Although Brooklyn has the most number of total shootings, it is second to Bronx in crime rate because of its large population. Notably Manhattan has a low crime rate, comparable to the lowest crime rate boroughs Queens, and Staten Island.
```{r, plot_crime_rate, warning=FALSE, message=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)

# Create a line chart for crime rates by borough
crime_rate_plot <- ggplot(crime_rates, aes(x = reorder(BORO, -crime_rate_per_100k), y = crime_rate_per_100k, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Crime Rates by Borough", x = "Borough", y = "Crime Rate per 100,000") +
  theme_minimal()

# Print the plot
print(crime_rate_plot)
```




