---
title: "Covid 19 Statistical Analysis"
author: "H. Justino"
date: '2024-12-07'
output:
  html_document: default
  pdf_document: default
---


## I want to understand if the deaths and cases of Covid 19 are consistent across States
## Note: I will provide a link to the github repo because I needed an absolute path to the datasets. Results will be reproducible.
#### Github Link: https://github.com/hermanjustino/Msc.-DS-Assignment-Repo/tree/main/Covid_19_Data_Analysis

#### Data is coming from Johns Hopkins University
```{r, import_data, message=FALSE, echo=FALSE, warning=FALSE}
global_cases <- read.csv("../datasets/time_series_covid19_confirmed_global.csv")
global_deaths <- read.csv("../datasets/time_series_covid19_deaths_global.csv")
us_cases <- read.csv("../datasets/time_series_covid19_confirmed_US.csv")
us_deaths <- read.csv("../datasets/time_series_covid19_deaths_US.csv")
```

### Tidy global cases

### These are US cases. It will include where the case was recorded, whether or not it was fatal.
```{r, tidy_global_cases, echo=FALSE, warning=FALSE}
library(tidyselect)
library(tidyr)
library(dplyr)
global_cases <- global_cases %>% 
  pivot_longer(cols=-c(`Province.State`,
                      `Country.Region`, Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat, Long))
```

## tidy global deaths
```{r, tidy_global_deaths, echo=FALSE, warning=FALSE}
library(tidyselect)
library(tidyr)
library(dplyr)
global_deaths <- global_deaths %>% 
  pivot_longer(cols=-c(`Province.State`,
                      `Country.Region`, Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))
```


#### First few rows of global cases. Country, status
```{r, inspect_date_column, echo=FALSE, warning=FALSE}
# Inspect the date column
head(global_cases$date)
```

## Join Cases with deaths
```{r, join_global_tables, echo=FALSE, warning=FALSE}
# Load necessary libraries
library(dplyr)
library(lubridate)

# Join Cases with deaths
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country.Region`,
         Province_State = `Province.State`) %>%
  mutate(date = gsub("^X", "", date),
         date = mdy(date))
```


### Tidy Us Cases
```{r, tidy_us_cases, echo=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(tidyr)

us_cases <- us_cases %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "date",
               values_to = "cases") %>%
  mutate(date = gsub("^X", "", date),
         date = mdy(date)) %>%
  select(Admin2:cases) %>%
  select(-c(Lat, Long_))
```

### Tidy Us Deaths
```{r, tidy_us_deaths, echo=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(tidyr)

us_deaths <- us_deaths %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "date",
               values_to = "deaths") %>%
  mutate(date = gsub("^X", "", date),
         date = mdy(date)) %>%
  select(Admin2:deaths) %>%
  select(-c(Lat, Long_))
```

### Join us tables
```{r, join_us_tables, echo=FALSE, warning=FALSE}
us <- us_cases %>%
  full_join(us_deaths)

```

### Filter us so only days with cases are visible
```{r, positive_filter_us}
us <- us %>% filter(cases > 0)

us
```

### Combine global

```{r, combine_global, echo=FALSE, warning=FALSE}
global <- global %>%
  unite(Combined_Key,
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```

### Filter global so only days with cases are visible
```{r, positive_filter_global, echo=FALSE, warning=FALSE}
global <- global %>% filter(cases > 0)
```

### Format UID 
```{r, format_uid, echo=FALSE, warning=FALSE}
uid <- read.csv("../datasets/UID_ISO_FIPS_LookUp_Table.csv") %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```

### Join global and uid
```{r, join_global_uid, echo=FALSE, warning=FALSE}
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, Population,
         Combined_Key)
```

### Summary
```{r, summary, echo=FALSE, warning=FALSE}
summary(us)
summary(global)
```

### US By State
```{r, us_by_state, echo=FALSE, warning=FALSE}
us_by_state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarise(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
```

### US totals
```{r, us_by_totals, echo=FALSE, warning=FALSE}
us_totals <- us_by_state %>%
  group_by(Country_Region, date) %>%
  summarise(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
```

### Filter US totals
```{r, filter_us_totals, echo=FALSE, warning=FALSE}
library(ggplot2)
library(scales)

us_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(colour = "cases")) +
  geom_point(aes(colour = "cases")) +
  geom_line(aes(y = deaths, colour = "deaths")) +
  geom_point(aes(y = deaths, colour = "deaths")) +
  scale_y_log10(labels = scales::label_number()) +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Covid 19 in US", y = NULL)
```

### Cases By State
```{r, cases_by_state, echo=FALSE, warning=FALSE}
library(stringr)
state <- "New York"

us_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(colour = "cases")) +
  geom_point(aes(colour = "cases")) +
  geom_line(aes(y = deaths, colour = "deaths")) +
  geom_point(aes(y = deaths, colour = "deaths")) +
  scale_y_log10(labels = scales::label_number()) +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("Covid 19 in ", state), y = NULL)

```

### New Cases
```{r, new_cases, echo=FALSE, warning=FALSE}
us_by_state <- us_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
us_totals <- us_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
```

### Graph New Cases
```{r, graph_new_cases, echo=FALSE, warning=FALSE}
us_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(colour = "new_cases")) +
  geom_point(aes(colour = "new_cases")) +
  geom_line(aes(y = new_deaths, colour = "new_deaths")) +
  geom_point(aes(y = new_deaths, colour = "new_deaths")) +
  scale_y_log10(labels = scales::label_number()) +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "New Covid 19 cases in US", y = NULL)
```

### Highest case rate in us
```{r, highest_case_rate_us, echo=FALSE, warning=FALSE}
us_state_totals <- us_by_state %>%
  group_by(Province_State) %>%
  summarise(deaths = max(deaths), cases = max(cases),
            Population = max(Population),
            cases_per_thou = 1000* cases/Population,
            deaths_per_thou = 1000* deaths/Population) %>%
  filter(cases > 0, Population > 0, deaths > 0)
```


###Highest Rate
```{r, highest_case_rate, echo=FALSE, warning=FALSE}
us_state_totals %>%
  slice_max(deaths_per_thou, n = 10)
```

### Lowest Rate
```{r, lowest_case_rate, warning=FALSE, echo=FALSE, warning=FALSE}
us_state_totals %>%
  slice_min(deaths_per_thou, n = 10)
```

### Model deaths vs cases
```{r, model_cases_deaths, warning=FALSE, echo=FALSE, warning=FALSE}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = us_state_totals)
summary(mod)
```

### Prediction vs actual on plot
```{r, plot_prediction, warning=FALSE, echo=FALSE, warning=FALSE}
us_state_totals %>% mutate(pred = predict(mod))
```

### Create prediction table
```{r, create_prediction_table, warning=FALSE, echo=FALSE, warning=FALSE}
us_state_totals_predictions <- us_state_totals %>% 
  mutate(pred = predict(mod))
```

### Plot prediction vs actual
```{r, plot_prediction_table, warning=FALSE, echo=FALSE, warning=FALSE}
us_state_totals_predictions %>% ggplot() + 
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

# Conlusion

#### This is a great dataset because it offers endless possiblities for manipulation and analysis. One source of bias I could forsee is in reporting. As data scientists, we can only analyze the data that is provided to us. If different places are reporting cases differently, it would effect our results.